Fast and adaptive cointegration based model for forecasting high frequency financial time series

### 2016-07-06 articles-conferences/iscef2016/src/results.tex
Final manuscript

\section{\uppercase{Experimental results}}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from Dukascopy~\cite{Dukascopy2014}, a free
database which gives access to the Swiss Foreign Exchange marketplace.

@@

Before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$. 
MacKinnon \cite{mackinnon2010} critical values for rejection of hypothesis of a unit root are: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
\begin{table*}[ht]
\label{tab:adf}
\centering
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level} \\
\multicolumn{6}{l}{ \textsuperscript{**} Indicates significance at 5\% level} \\
\multicolumn{6}{l}{ \textsuperscript{***} Indicates significance at 10\% level} \\
\multicolumn{6}{l}{MacKinnon critical values for rejection of hypothesis of a unit root are:}\\
\multicolumn{6}{l}{ -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}\\
\multicolumn{6}{l}{ADF($d$) Augmented Dickey-Fuller test with lag $d$} 
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\end{table*}
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root test in their level form considering different lags, but they rejected it with their first differences. This means that all of
them are I(1) time series and we are allowed to use VECM and therefore our proposed AVECM.

@@

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of data (from the 12th to the 15th of August 2014).
For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.

@@

The Johansen method is already programmed in the Python Statsmodels
library~\cite{seabold2010} and the parallel implementation was developed using
MPI in Python. 
We chose MPI because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
In order to compare serial and parallel execution times in AVECM, we set parameter $it=100$ in algorithm\ref{alg:AVECM}.

The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{img/extimes}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}

>>>>

\section{Experimental results}
\label{sec:results}
\subsection{Data} \label{sec:unitroot}
AVECM tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. This data was collected from
Dukascopy \citep{Dukascopy2014}, a free database which gives access to the Swiss
Foreign Exchange marketplace.

@@

Before running the tests, we firstly checked whether the time series were I(1)
using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.  MacKinnon
\citep{mackinnon2010} critical values for rejection of hypothesis of a unit root
are: -2.62 (1\%), 1.94 (5\%) and 1.62 (10\%).
\begin{table*}[ht]
\label{tab:adf}
\centering
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level} \\
\multicolumn{6}{l}{ \textsuperscript{**} Indicates significance at 5\% level} \\
\multicolumn{6}{l}{ \textsuperscript{***} Indicates significance at 10\% level} \\
\multicolumn{6}{l}{MacKinnon critical values for rejection of hypothesis of a unit root are:}\\
\multicolumn{6}{l}{ -2.62 (1\%), -1.94 (5\%) and -1.62 (10\%)}\\
\multicolumn{6}{l}{ADF($d$) Augmented Dickey-Fuller test with lag $d$} 
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-seconds
frequency.}
\end{table*}
Table~\ref{tab:adf} shows that all currency rates cannot reject the unit root
test in their level form considering different lags, but they rejected it with
their first differences. This means that all of them are I(1) time series and we
are allowed to use VECM and therefore our proposed AVECM.

@@

Algorithms AVECM, ARIMA and the naive random walk were tested using four days of
data (from the 12th to the 15th of August 2014).  For AVECM we considered
different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10,
50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and
$p$. Possible values for $L$ were in the interval $[2,14]$ hours and $p$ can
have values in between $[1,5]$. Best AVECM performance was compared against
ARIMA and the random walk model.  Table~\ref{tab:stats} shows the out-of-sample
performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of
both measures we found that AVECM is superior to ARIMA and the naive random walk
model. We also included the p-value that proves that the difference in the MSE
is significant at the 99\% significance level in three of the four currency
rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and
ARIMA are superior to the naive random walk model and that our proposal is also
superior to ARIMA.

@@

The Johansen method is already programmed in the Python Statsmodels
library \citep{seabold2010} and the parallel implementation was developed using
MPI in Python.  We chose MPI because it allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters or on
local computers.  We tested our proposal in a cluster with 2 servers Xeon
E5-2667 (2.90GHz) of 24 cores each (48 cores in total) and 24GB RAM.  In order
to compare serial and parallel execution times in AVECM, we set parameter
$it=100$ in algorithm\ref{alg:AVECM}.

The $L$ parameter was always chosen between 2 and 14 hours and $p$ always took
values between 1 and 5. Parameter $nparams$ represents the number of pairs
($L$,$p$) used to maximise the percentage of cointegration.

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=0.75\textwidth]{img/Fig3}
  \caption{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}
